% ######################################################################################################################
%         Conclusion and Future Directions
% ######################################################################################################################

% \chapter{Conclusion and Outlook}
\chapter{Conclusion and Future Directions}
\label{ch:ConclusionFutureDirections}

\paperbox{
    This chapter mostly contains original contributions by Lucas Czech written for this thesis.
    Some of the text is however derived from the concluding sections of the peer-reviewed open-access publication:
}{\paperpppp}{
%     In particular, we adapted the summaries and future directions of our methods to the context of this chapter.
%     All text in this chapter was written by Lucas Czech.
    The respective text was originally written by Lucas Czech, and modified here to fit into the context of this chapter.
}

In this thesis, we made several contributions to the field of computational phylogenetics,
and phylogenetic placement in particular.
% The methods were already described in two peer-reviewed publications \cite{Czech2018,Czech2018a},
% and the respective data and scripts were made available at \url{http://github.com/lczech/placement-methods-paper}.
We have already described the methods in two peer-reviewed publications \cite{Czech2018-phat,Czech2019-analysis},
and made the respective data and scripts available at \url{http://github.com/lczech/placement-methods-paper}.
The implementation is available in our \toolname{genesis} library (\url{https://github.com/lczech/genesis});
we furthermore offer a ready-to-use command line interface for the presented methods
via our \toolname{gappa} tool (\url{https://github.com/lczech/gappa}).
Both these tools are described in \appref{ch:PipelineImplementation}
as well as our application note \cite{Czech2019-genesis-gappa}.
This chapter briefly summarizes our contributions and discusses potential future developments
and open research questions in the field of phylogenetic placement.

Having been introduced between 2009 and 2010 \cite{Berger2009,Matsen2010a},
phylogenetic placement and its downstream analysis methods are still relatively recent tools,
but are becoming more and more popular in the research community.
Their advantages compared to other metagenomic analysis methods
include improved interpretability and visualizability of results,
at the cost of increased methodological and computational complexity.
There are hence still obstacles to overcome before phylogeny-aware methods (such as phylogenetic placement)
can reach maturity and widespread use in the research community:
% generally less prevalent than simpler analysis methods,
% such as those based on sequence similarity.

\begin{itemize}
    \item \textbf{Usability:}
          Current software pipelines for phylogenetic placement are mostly build from separate programs and tools
          that are not well integrated with each other, and only interact via intermediate files and scripts.
          This induces labor costs for first-time users of these tools,
          for instance, because of the manual compilation and setup of the software,
          as well as the time required for ad-hoc scripting solutions to connect different steps of the pipeline.
    \item \textbf{Scalability:}
          The amount of available sequence data is growing exponentially (see \secref{ch:Introduction:sec:Motivation}),
          currently doubling roughly every seven months \cite{Stephens2015}.
          Phylogeny-aware analysis methods for sequence data are generally more computationally intense than, for instance,
          methods that are based on sequence similarity.
          The speed and scalability of these tools thus
          need to stay on a par with the amount of sequencing data being produced.
    \item \textbf{Analysis Methods and Research Questions:}
          Downstream analysis methods that take phylogenetic information into account
          are not yet as versatile and mature as classical data analysis methods.
          Hence, researchers might refrain from using phylogenetic placement and resort to more established methods.
          There are however numerous research questions that might benefit from analyses using phylogeny-aware methods,
          for example, in disciplines such as bio-geography or medicine.
%           Methods are missing maybe because of the complexity.
\end{itemize}

In the following, we discuss our contributions to the field with respect to these obstacles,
and describe potential future research directions.

% ######################################################################################################################
%         Usability and Scalability
% ######################################################################################################################

\section{Usability and Scalability}
% \paragraph{Usability and Scalability}
\label{ch:ConclusionOutlook:sec:UsabilityScalability}

In this section, we discuss our contribution to software development for analyzing phylogenetic data,
as well future directions for such software in general, and our software in particular.

\paragraph{Contribution}
\label{ch:ConclusionOutlook:sec:UsabilityScalability:par:Contribution}

In \chpref{ch:AutomaticTrees}, we presented an approach to automatically obtain
reference trees for phylogenetic placement called \emph{\acfp{PhAT}} using large reference sequence databases.
We showed that \acp{PhAT} are valuable and sufficiently accurate trees for
conducting phylogenetic placement and taxonomic assignment of metagenomic sequences.
They thus help with the labor-intense and potentially error-prone process of collecting suitable reference sequences,
and might even replace such ``manual'' trees in some applications.
% Hence, an interesting future direction might be to extend and evaluate them for other purposes and use cases such as ???
In the same chapter, we presented a multi-level placement approach as well as our pre-processing pipeline
to enable and accelerate phylogenetic placement
of large, diverse datasets with hundreds to thousands of environmental sequence samples.
% All these methods simplify and accelerate the processes for researchers
% who want to apply phylogenetic placement and respective downstream analyses.

\paragraph{General Future Development}
\label{ch:ConclusionOutlook:sec:UsabilityScalability:par:FutureDevelopment}

In order to further facilitate the use of phylogenetic placement methods,
without the need for complex setups and scripts, online services for quickly testing standard methods might be worth to offer.
For example, uploaded query sequences could be placed on our four \acp{PhAT} used in the evaluation
(\secref{ch:AutomaticTrees:sec:Evaluation:sub:ReferenceTreeSetup:par:SequenceSelection}),
or on custom reference trees.
These services could further offer standard analysis methods such as Edge~PCA and Squash Clustering
(\secref{ch:Foundations:sec:PhylogeneticPlacement:sub:ExistingMethods}),
or the methods that we presented in this thesis (Chapters \ref{ch:Visualization}--\ref{ch:Factorization}).

For more advanced or large-scale studies, a custom setup is mostly inevitable,
because users need to scale up the processing using computer clusters as well as customize the workflow.
% State-of-the-art implementations of phylogenetic placement, such as our \toolname{EPA-ng} \cite{Barbera2018},
% are able to place billions of sequences withing a few hours on computer clusters.
% There are however other steps in typical pipelines that do not yet scale and hence form a bottleneck.
% For example, \toolname{PaPaRa} \cite{Berger2011a,Berger2012},
% a phylogeny-aware tool for aligning query sequences against a reference alignment,
% might benefit from a re-implementation.
Such studies might benefit from offering well-defined integrated pipelines for the basic steps of phylogenetic placement,
with a simple setup for all common functionality, for instance by using platforms such as
\toolname{Galaxy} \cite{Afgan2018} or \toolname{Conda}/\toolname{Bioconda} \cite{Gruning2018}.
Tools and pipelines that integrate different steps of the analysis furthermore allow for more efficient implementations,
as they can make better use of computing infrastructure,
e.\,g., via parallel programming interfaces such as \texttt{\acs{MPI}},
and distribute computations with less overhead for intermediate files and bookkeeping between pipeline steps.

% can be implemented in integrated ways for computer clusters.
% which will be necessary because of the large amounts of data that are coming our way

Moreover, to achieve a tighter integration of the tools in a pipeline,
and to incentivize experimentation with existing methods and development of novel methods,
file formats need to be flexible and extensible.
For example, the \fileformat{jplace} file standard for phylogenetic placement \cite{Matsen2012}
is based on the \fileformat{JSON} format \cite{JsonMemo,JsonStandard}, and hence easily extensible.
The standard however currently (as of \fileformat{jplace} version 3) lacks support for multiple samples per file,
and additional per-sample or even per-query annotations and other meta-data.
A similar issue applies to file formats for phylogenetic trees, as we showed in \citeay{Czech2017-tree-viewers}.
In particular, the \fileformat{Newick} format is often extended ad-hoc for specific needs
(this is even done within the \fileformat{jplace} format itself),
with no standardized support from downstream tools or tree viewers.
% leading to mis-interpretations of and errors in results \cite{Czech2017-tree-viewers}.
This leads to ambiguities, misinterpretations, and errors in study results
that could be avoided by creating proper standards for these data types \cite{Czech2017-tree-viewers}.
% Creating proper standards could thus also help to avoid ambiguities and misinterpretations of results.

% before researchers actually tune existing methods to their particular needs
% and start extending and experimenting with existing approaches.
% hinders rapid testing of methods,

\paragraph{Future Development of our Software}
\label{ch:ConclusionOutlook:sec:UsabilityScalability:par:FutureOfOurSoftware}

With respect to the methods and tools presented in this thesis,
we plan to extend our \toolname{gappa} tool \cite{Czech2019-genesis-gappa} to include additional existing analysis methods,
such as placement-based diversity measures \cite{Matsen2010} and UniFrac calculations \cite{Lozupone2005},
and additional pre- and post-processing functions, such as filtering, merging and manipulating placement files.
In the long term, we also intend to re-implement all functionality offered by the
\toolname{guppy} tool from the \toolname{pplacer} suite \cite{Matsen2010},
to turn \toolname{gappa} into a more efficient and scalable replacement for \toolname{guppy}.
Moreover, there is a \toolname{Conda}/\toolname{Bioconda} \cite{Gruning2018} recipe
for installing \toolname{gappa} \cite{Douglas2018-gappa}.
The recipe was created as part of the \toolname{PICRUSt2} pipeline \cite{Douglas2019},
which internally relies on \toolname{gappa} for some of its steps.
The recipe hence is not maintained by us;
we however plan to monitor it in order to ensure its usability.

Furthermore, we are planning to extend the \toolname{genesis} library \cite{Czech2019-genesis-gappa},
which is written in \texttt{C++11}, to also offer \acs{API} bindings for \texttt{Python}.
Hence, \toolname{genesis} could combine the advantages of both programming languages:
For standard functions such as file processing and most internal computations,
it can use our highly efficient \texttt{C++} implementations,
which relies on multi-threading via \texttt{OpenMP} (Open Multi-Processing)
where appropriate to allow usage of the available compute cores.
The modular and (hopefully) clear \acs{API} of \toolname{genesis} further allows it to be used in production code,
% as evident by its use in \toolname{gappa} and the quartet-based tree distance measure that we contributed to \cite{Zhou2017}.
as evident by its use in several of our publications
\cite{Mahe2017,Barbera2018,Zhou2017,Czech2018-phat,Czech2019-analysis,Czech2019-genesis-gappa}.
At the same time, ad-hoc solutions for pipeline tasks such as file conversions or extraction of certain data
could be implemented as \texttt{Python} scripts.
This also facilitates to use \toolname{genesis} for experimentation and rapid prototyping of novel ideas and methods.

% ######################################################################################################################
%         Analysis Methods
% ######################################################################################################################

\section{Analysis Methods}
\label{ch:ConclusionOutlook:sec:AnalysisMethods}

Downstream methods for analyzing metagenomic data in a phylogenetic context,
and in paricular, methods based on phylogenetic placements,
are not yes as plentiful as standard sequence-based methods.
We contributed several novel ideas and adaptations of existing concepts,
which we summarize in the following.
Then, we present some ideas for adapting existing machine learning techniques to phylogenetic data.

\paragraph{Contribution}
\label{ch:ConclusionOutlook:sec:AnalysisMethods:par:Contribution}

In \chpref{ch:Visualization}, we described methods for visualizing the phylogenetic placement of large metagenomic datasets.
The methods allow to detect differences between samples (\emph{Edge Dispersion}),
as well as correlations with per-sample meta-data (\emph{Edge Correlation}),
and are thus intended for similar use cases as the established Edge PCA \cite{Matsen2011a}.
However, our novel methods directly visualize important features of the samples (and their meta-data)
on the underlying reference tree.
This allows for interpreting them in a phylogenetic context.
Furthermore, in \chpref{ch:Clustering}, we introduced clustering methods for metagenomic samples
(\emph{Phylogenetic $k$-means} and \emph{Imbalance $k$-means}),
which serve a similar purpose as Squash Clustering \cite{Matsen2011a},
but are better suited for larger datasets due to their fixed output size.
% The presented methods can also be combined, allowing

In \chpref{ch:Balances}, we introduced an adaptation of the Phylogenetic ILR transformation and balances \cite{Silverman2017}
to phylogenetic placements.
Our adaptation uses a fixed reference tree (instead of a tree inferred from the OTUs present in the data),
and allows to place sequences along the branches of the tree (instead of only at its tips).
% Balances are conceptually similar to edge imbalances, which are used in Edge PCA \cite{Matsen2011a} and in our Edge Correlation.
% and are an approach to contend with the compositional nature of metagenomic data.
% A further approach to contend with the compositional nature of metagenomic data are
% the node-based balances using the isometric log-ratio of OTU abundances \cite{Silverman2017,Washburne2017a}
% that we discussed in \secref{ch:Foundations:sec:PhylogeneticPlacement:sub:PlacementProcessing:par:EdgeImbalances}.
% methods related to edge imbalances
% For future research, these methods could be adapted to phylogenetic placement data.
% To this end, they need to be extended from abundances ``placed'' on the tips of the OTU tree
% to masses placed along the branches of a reference tree.
As balances are a transformation that yields orthogonal components, % (one for each node or edge of the tree),
issues like the normalization of compositional data
(\secref{ch:Foundations:sec:PhylogeneticPlacement:sub:PlacementProcessing:par:Normalization}) do not arise.
% Applying these methods to placements instead of OTUs allows for more detailed analyses.
% Furthermore, using a fixed reference tree %instead of one inferred from the OTUs present in a set of samples
% allows comparative studies across datasets.
As we showed, with samples being represented as a vector of balances,
many standard tools for visualization, ordination, and clustering of data in the Euclidean space
can be readily applied to phylogenetic placement data.
% Lastly, visualizations similar to Edge Correlation (\secref{ch:Visualization:sec:Methods:sub:EdgeCorrelation})
% can be achieved with such data, by relating the balance per node with meta-data features.
% Such a visualization would highlight nodes that exhibit a strong correlation
% between changes in the balance of their subtree with environmental variables,
% while solving many of the issues of compositional data that our approach might suffer from.

In \chpref{ch:Factorization}, we presented an adaptation of Phylofactorization \cite{Washburne2017a} to phylogenetic placements,
which we call \emph{Placement-Factorization}.
These methods are able to identify edges of the tree that exhibit predictable changes in species abundances
based on per-sample meta-data features.
This allows to understand evolutionary and ecological patterns (e.\,g., abundances in certain clades of the tree)
that are driven by environmental factors (e.\,g., pH value of the soil).
The output and figures produced by these methods allow for a detailled interpretation of the results,
and allow to simultaneously understand many different aspects of the data.
As outlined in the chapter, the capabilities of these methods have just started being explored.
We hence see large potential for futher developing Phylofactorization and Placement-Factorization.

% \todo{ maybe add this instead as an outlook idea: } \verb|https://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-018-0584-3|
% \todo{from Justin, 11/21/18, 15:49:
% One thought: Have you considered building a poisson or multinomial model and letting the PhILR transform be the link function?
% We have a new paper using this strategy.  It works quite well. It relies on either a prior (bayesian) or regularization
% (frequentist-ish) to ensure that the logs don't push estimates to negative infinity. In some senses it works like model
% based zero-imputation. We find it works much better than pseudo-counts. Using this strategy the count model takes
% the place of the taxa weights (which are then not needed/used).
% }

\paragraph{Classical Machine Learning Approaches}
\label{ch:ConclusionOutlook:sec:AnalysisMethods:par:MachineLearningApproaches}

The novel methods that we introduced in Chapters \ref{ch:Visualization}--\ref{ch:Factorization}
extend the portfolio of available analysis tools for phylogenetic placement.
Additionally, there exist several approaches from the fields of machine learning and data mining
that could be useful to adapt to phylogenetic placement data.
It might also be worth to develop integrative methods that can incorporate heterogeneous features
such as phylogenetic data and different types of meta-data in a combined analysis \cite{Mariette2018}.
While Placement-Factorization already is a first step in this direction,
there is more potential for fully integrated data analysis methods.
% e.\,g., using unsupervised machine learning methods \cite{Mariette2018}.
The challenge for adapting existing methods usually consists in making the methods phylogeny-aware,
for example by having them operate on mass distributions on trees
(including respective distance measures such as the KR distance) rather than on $\mathbb{R}^d$ vectors
that are typically used in machine learning algorithms.

As we showed, %even without the use of such balance approaches,
\emph{unsupervised} methods (such as $k$-means clustering) can be extended to phylogenetic placement data,
and are thus valuable tools for metagenomic data exploration.
It might therefore be interesting to also adapt other types of unsupervised machine learning methods to such data,
for instance, different clustering \cite{Kriegel2011} or dimensionality reduction \cite{VanDerMaaten2008} techniques,
or visualization \cite{Laczny2014} and anomaly detection approaches.

The adaptation of \emph{supervised} machine learning approaches to phylogenetic data might however be more complicated.
In recent years, several approaches have been proposed for metagenomic sequence data,
using large-scale machine learning \cite{Vervier2015,Soueidan2015}
and deep learning \cite{Angermueller2016,Fiannaca2018} techniques.
These approaches consider each sequence to be a data point, meaning that enough training data are available.
They are applied for tasks such as OTU clustering, sequence classification, taxonomic assignment, and gene prediction.
In this thesis however, we are interested in entire metagenomic samples (instead of single sequences)
as the data points to be considered.
Current datasets are often not large enough (in number of distinct samples)
to allow for robust learning at this level of granularity without over-fitting the training data \cite{Angermueller2016}.
There are however some recent approaches in this direction, which focus on comparative studies \cite{Soueidan2015}.
They are mainly based on features such as
abundances and presence/absence patters of sequences (or \acsp{OTU}) within the samples \cite{Pasolli2016},
or use so-called $k$-mers (subword occurrences within sequences) as features \cite{Asgari2015}.
These features are simple enough to allow for training with current dataset sizes.
However, given the growth rate of metagenomic sequence data,
sufficient training data for more detailed analyses are likely to become available in the future.
Hence, incorporating phylogenetic information in supervised machine learning methods might yield
improved accuracy as well as better interpretability.
This could be used for tasks such as classification of samples,
% (e.\,g., prediction of meta-data features such as health status with respect to a disease)
and prediction and regression with respect to per-sample meta-data features.
Placement-based balances and phylogenetic factorization
are potential ways to approach this idea \cite{Silverman2017,Washburne2017a},
as indicated by the results of this thesis.

\paragraph{Neural Networks and Deep Learning Approaches}
\label{ch:ConclusionOutlook:sec:AnalysisMethods:par:DeepLearningApproaches}

Notable supervised approaches that are interesting and promising for problems in the life sciences
are \emph{neural networks}, and \emph{deep learning} in particular \cite{Skansi2018}.
Such methods are, in principle, well suited for the complex and high-dimensional data produced in these domains.
Initial approaches have recently been described
for a variety of use cases in biology and medicine \cite{Mamoshina2016,Min2017,Ching2018}.
There are however some general issues that need to be solved when employing deep learning.
Firstly, similar to other supervised methods, the amount of available (labeled) data in current biological studies
is often too limited for training deep neural networks \cite{Min2017,Ching2018}.
Secondly, deep learning is often considered as a ``black box'' that lacks interpretability and testability;
there are however recent approaches to alleviate this \cite{Min2017,Reiman2018,Wainberg2018}.
% how users should interpret these models to make testable hypotheses about the system under study remains an open challenge.
Future data collection and method development efforts will likely be able to overcome these obstacles \cite{Wainberg2018}.
Then, deep learning approaches that take phylogenetic information into account
could represent viable tools for analyzing and understanding metagenomic data.

There have also been first attempts that directly use tree structures and phylogenetic trees in neural networks.
For example, the architecture of a neural network can be modeled
according to a given graph structure \cite{Scarselli2009,Bronstein2017}.
In such models, data on the nodes of the graph is propagated across the graph during the training of the network.
This allows to exploit the additional information encoded in the graph.
This could be used for phylogenetic trees as well, using the tree as a fixed graph,
and the distribution of phylogenetic placements as the data that the system is trained on.
A very recent and promising idea is to embed the evolutionary information of phylogenetic trees into the Euclidean space,
and then use convolutional neural networks (CNNs) on the resulting matrices
to analyze the underlying data \cite{Fioravanti2018,Reiman2017,Reiman2018}.
These methods are able to accurately predict meta-data features of a set of metagenomic samples,
given the per-sample taxonomic profile.
In detail, these methods infer a tree from the OTUs present in the dataset,
and use the per-sample OTU abundances to train the network.
In consequence, the underlying tree differs in each study, which hinders the use of pre-trained networks,
for example in medical applications, where data of new patients needs to be classified.
Instead, one could easily embed a fixed reference tree in a similar manner,
and use phylogenetic placement distributions on the tree instead of abundances for the training.
% These recent methods can be apply with good prospects to metagenomic data, and to phylogenetic placement in particular.

\paragraph{Incorporating Uncertainty}
\label{ch:ConclusionOutlook:sec:AnalysisMethods:par:Uncertainty}

% From Factorization Chapter:
% In an ideal world with a known phylogeny of all of life, alternative OTU clusterings and alternative trees
% would simply collapse nodes at different depths (pers.~comm. with A.~Washburne on 2019-03-01).
% Unfortunately, real world data, and particularly different OTU clustering methods and tree inference methods,
% will yield discordant trees.
% The influence of uncertainty in the phylogeny is further discussed in \citeay{Washburne2019}.

Phylogenetic placements already incorporate a measure of uncertainty for each query sequence
in form of the likelihood weight ratio (LWR) of each placement position
(\secref{ch:Foundations:sec:PhylogeneticPlacement:sub:PipelineAndComputation:par:ComputingQuerySequence}).
They however assume that the given fixed reference tree on which the placement is conducted is correct.
% In all methods that we presented here, the reference tree is generally assumed to be fixed.
As discussed in \secref{ch:Foundations:sec:TreeOfLife:sub:PhylogeneticTrees},
the ``true'' tree of life is unknowable \cite{Haber2005}.
In an ideal world, alternative OTU clusterings and alternative trees would simply collapse nodes at different depths
(pers.~comm. with A.~Washburne on 2019-03-01, c.\,f. \secref{ch:Factorization:sec:Evaluation:sub:OralFecalHMPDataset:par:Comparison}).
Due to uncertainty in both the OTU clustering and the tree inference however, errors are introduced.

Measures such as \emph{bootstrap support values} \cite{Efron1979,Felsenstein1985a,Soltis2003}
and methods such as \emph{Bayesian inference} (\secref{ch:Foundations:sec:TreeOfLife:sub:TreeInference})
can help to assess the uncertainty introduced by these errors.
The possibility to incorporate the uncertainty of the tree inference
has already been discussed in the context of Phylofactorization \cite{Washburne2019}:
the more certain the scaffold provided by the phylogeny,
the more certain the inferences about a clade obtain from the factorization.
For future development, it might be informative to incorporate uncertainty in the analyses methods presented here as well.

% ######################################################################################################################
%         Research Questions
% ######################################################################################################################

\section{Research Questions}
\label{ch:ConclusionOutlook:sec:ResearchQuestions}

% future of phylogeny-aware methods, in particular of phylogenetic placement:
% concepts such as deep learning and machine learning in general
% There is however also potential in the development of more concrete methods for particular applications and use cases.
% Machine learning techniques could for example be used to this end,
% as tools for developing methods for the following research questions
% and provide ample prospects for future development.
% Using such tools and methods can help answer important research questions in biology,
% This can help to answer important research questions

The general concepts and techniques described above %are of current interest in the research community, and
can be employed for developing more specialized methods to help answering novel research questions.
% for example in fields such as biogeography, ecology, and human health.
In the following, we present some ideas for future applications and use cases in metagenomics,
to which phylogeny-aware methods, and phylogenetic placement in particular, could be adapted
by using these techniques.

\paragraph{Contribution}
\label{ch:ConclusionOutlook:sec:ResearchQuestions:par:Contribution}

% a first step to answer known questions in a new phylogeny-aware manner ...
As mentioned in \secref{ch:Introduction:sec:ContributionOverview}, we are currently working on \toolname{scrapp},
which stands for ``Species Counting on Reference trees viA Phylogenetic Placements''.
It is a tool for estimating the per-branch species diversity in metagenomic samples on a given reference tree,
and is hence useful for discovering and describing novel species in microbial data.
Thus, \toolname{scrapp} adapts a common question in biology (species counting and diversity estimation),
and brings it into a phylogenetic context,
where results can be visualized and interpreted using the additional information of the reference tree.
It is a pipeline that combines several tools developed in our lab, such as
\toolname{EPA-ng} \cite{Barbera2018}, \toolname{ParGenes} \cite{Morel2019}, and \toolname{mPTP} \cite{Kapli2017},
and uses \toolname{genesis} \cite{Czech2019-genesis-gappa} for file conversions
and other intermediate tasks between pipeline steps.
It is hence a first step in the direction of integrated pipelines as explained above.

% Apart from species diversity, there are
% Estimating species diversity is an important task for metagenomic studies in general.
% and hence in demand by the research community.

A related question is that of estimating the microbial community composition in a metagenomic sample,
for example in the form of per-taxon relative abundances and taxonomic assignments \cite{Lindgreen2016}.
We explained the compositional nature of phylogenetic placements in
\secref{ch:Foundations:sec:PhylogeneticPlacement:sub:PlacementProcessing:par:EdgeMasses},
and described our ad-hoc implementation for taxonomic assignment based on phylogenetic placement in %the \toolname{gappa} tool in
\secref{ch:AutomaticTrees:sec:Evaluation:sub:TaxonomicAssignmentProfiling}.
In order complement methods specifically developed for taxonomic assignment
\cite{Lindgreen2016,Sczyrba2017,Almeida2018},
our implementation could be extended and refined,
and hence become a tool that uses phylogenetic context to answer questions of community composition.

% There are furthermore some specific fields in which we see an opportunity for phylogeny-aware methods,
% for example biogeography, ecology, and medicine. %human health.

\paragraph{Further Research Directions}
\label{ch:ConclusionOutlook:sec:ResearchQuestions:par:FurtherResearch}

Further fields in which we see an opportunity for developing phylogeny-aware methods are biogeography and ecology,
as they ask questions concerning the distribution of species across geographic space and through time,
as well as regarding their interactions with each other.

Research has been conducted on unravelling the bio-geographical patterns of microbial communities
\cite{HughesMartiny2006}, and there also exist evolutionary approaches \cite{Cox2016}.
This could, for instance, be extended to phylogenetic placements by combining them with geographical data:
The phylogenetic Kantorovich-Rubinstein (KR) distance between two metagenomic samples is a pairwise measure
(\secref{ch:Foundations:sec:PhylogeneticPlacement:sub:Distances:par:KR});
similarly, the geographic locations where the samples were collected yield pairwise distances between samples.
These measures could be used to infer microbial patterns that change with geographical distance.
This is similar to correlating the placement distribution with per-sample meta-data
as described in \secref{ch:Visualization:sec:Methods:sub:EdgeCorrelation},
but requires some other form of correlation measure.
% pairwise measures could be used for this.
We conducted initial tests in this direction, but unfortunately,
the available datasets were shown to not exhibit such biogeographical patterns \cite{Lentendu2018}.

Instead of location, time is also used for comparing metagenomic samples.
In such studies, the same location is sampled at different points in time,
thus allowing to investigate the dynamics, variations, and (potentially) seasonal periodic patterns
experienced in the microbial communities across time \cite{Costello2009,Faust2015}.
This is typically assessed using the taxonomic composition of the samples.
By extending such methods to phylogenetic placements,
the time-dependent patterns could be displayed on a phylogenetic tree (or a series of trees),
which again might provide additional information for interpreting and visualizing them.
One could for instance construct a time series (video) of a tree showing
how abundance patterns similar to \figref{fig:abundances_comparison:sub:branch_colors} vary over time.

% instead of assessing dependencies between species abundances and external factors (such as meta-data features, location, time),
Furthermore, in the field of ecology, the co-occurrence patterns between species
have been studied for numerous habitats and ecosystems \cite{Lima-Mendez2015,Villalobos2017,Zelezniak2015}.
These interactions are often visualized in so-called co-occurrence networks \cite{Chaffron2010,Faust2012},
which aggregate abundances at a high taxonomic level,
and show links that indicate co-presence and exclusion between the depicted taxa.
Unfortunately, even at high taxonomic levels, these networks are often extremely complex and hard to interpret,
while potentially omitting valuable interaction details at lower taxonomic levels.
Thus, it might be interesting to visualize such co-occurrences with respect to phylogenetic information.
This entails the challenge to simplify the visualization and only show relevant co-occurrences,
for example by highlighting branches of the reference tree that exhibit a high correlation with other branches.
% Such methods might then be

Throughout this thesis, we used the \acf{BV} \cite{Srinivasan2012} and the \acf{HMP} \cite{Huttenhower2012,Methe2012}
datasets as examples for health-related applications of our methods.
Many of the above research questions are also interesting from a medical point of view,
and can be readily applied to other human datasets for assessing health status.
In recent years, there have been studies that conduct phylogenetic analyses of widely researched diseases
such as HIV/AIDS \cite{Castro-Nallar2012,Brenner2013,Ratmann2017} and different types of cancer \cite{Brown2017a,Abbosh2017}.
Thus, we see a large potential in phylogeny-aware methods for diagnostics, and human health in general.

In conclusion, the development of novel methods that use phylogenetic information
is highly promising for many fields of research in biology, medicine, and the life sciences.
In particular, methods based on phylogenetic placement allow to process large metagenomic datasets,
and to visualize them in different ways using the underlying reference tree as an additional source
of information that aids interpretation of the results.
