% ######################################################################################################################
%         Publications
% ######################################################################################################################

\cleardoublepage

\chapter{List of Publications}
\label{ch:Publications}

here: all publications.
see also \secref{ch:Introduction:sec:ContributionOverview} for more details on the important major ones.

\todo{workshops and conferences, TA}

\todo{SA and DA?}

\todo{make sure to update the bioRxiv preprints to the actual publications! not only here, but everywhere! my papers, quartet, epa, etc}
\todo{same for epa ng paper! }

My Submissions for VIII ECOP-ISOP 2019 Joint meeting

55	Mahwash Jamy, Rachel Foster, Lucas Czech, Alexey Kozlov, Pierre Barbera, Alexandros Stamatakis, David Bass and Fabien Burki	Long metabarcoding of the eukaryotic rDNA operon to phylogenetically and taxonomically resolve environmental diversity

214	Cédric Berney, Frédéric Mahé, Nicolas Henry, Jeena Rajan, Lucas Czech, Pierre Barbera, Alexandros Stamatakis, Guy Cochrane and Colomban de Vargas	EukBank: a community resource to explore the Earth eukaryotic diversity

\todo{make sure that biorxiv pages are consistently in or out!}

\begin{enumerate}
    \item \textbf{L. Czech}, S. Berger, D. Krompaß, J. Zhang, P. Kapli, P. Pavlidis, and A. Stamatakis.
        Evolutionary Placement of Short Reads - Methods, Applications, and Visualization.
        Poster at \textit{EMBO/EMBL Symposium: A New Age of Discovery for Aquatic Microeukaryotes}, Heidelberg, Germany, January 2016,
        and at \textit{Hellenic Bioinformatics Conference (HBio) 2016}, Thessaloniki, Greece, November 2016.

    \item F. Mah{\'{e}}, C. de Vargas, D. Bass, \textbf{L. Czech}, A. Stamatakis, E. Lara, D. Singer, J. Mayor, J. Bunge,
        S. Sernaker, T. Siemensmeyer, I. Trautmann, S. Romac, C. Berney, A. Kozlov, E. A. D. Mitchell, C. V. W. Seppey,
        E. Egge, G. Lentendu, R. Wirth, G. Trueba, and M. Dunthorn.
        Parasites dominate hyperdiverse soil protist communities in Neotropical rainforests.
        \textit{Nature Ecology \& Evolution}, 1(4):0091, 2017. \cite{Mahe2017}

    \item \textbf{L. Czech}, J. Huerta-Cepas, and A. Stamatakis.
        A Critical Review on the Use of Support Values in Tree Viewers and Bioinformatics Toolkits.
        \textit{Molecular Biology and Evolution}, 17(4):383--384, 2017. \cite{Czech2017-tree-viewers}

    \item T. Flouri, J. Zhang, \textbf{L. Czech}, K. Kobert, and A. Stamatakis.
        An efficient approach to merging paired-end reads and the incorporation of uncertainties.
        In M. Elloumi, editor, \textit{Algorithms for Next-Generation Sequencing Data: Techniques, Approaches and Applications},
        chapter 13, pages 299--326. Springer International Publishing AG, 1 edition, 2017. \cite{Flouri2017}
%         Chapter in \textit{Algorithms for Next-Generation Sequencing Data: Techniques, Approaches and Applications}.
%         1st ed., M. Elloumi, Ed. Springer International Publishing AG, 2017, pp. 299--326. \cite{Flouri2017}

    \item P. Barbera, A. Kozlov, T. Flouri, D. Darriba, \textbf{L. Czech}, and A. Stamatakis.
        Massively Parallel Evolutionary Placement of Genetic Sequences.
        Poster at \textit{ISC 2017 PhD Symposium}, Frankfurt am Main, Germany, June 2017.

    \item C. Berney, A. Ciuprina, S. Bender, J. Brodie, V. Edgcomb, E. Kim, J. Rajan, L. W. Parfrey, S. Adl, S. Audic,
        D. Bass, D. A. Caron, G. Cochrane, \textbf{L. Czech}, M. Dunthorn, S. Geisen, F. O. Glöckner, F. Mah{\'{e}}, C. Quast,
        J. Z. Kaye, A. G. B. Simpson, A. Stamatakis, J. del Campo, P. Yilmaz, and C. de Vargas.
        UniEuk: Time to Speak a Common Language in Protistology!
        \textit{Journal of Eukaryotic Microbiology}, 38(1):42--49, 2017. \cite{Berney2017}

    \item X. Zhou, S. Lutteropp, \textbf{L. Czech}, A. Stamatakis, M. von Looz, and A. Rokas.
        Quartet-based computations of internode certainty provide accurate and robust measures of phylogenetic incongruence.
        \textit{bioRxiv}, 168526, 2017. \cite{Zhou2017}

    \item P. Barbera, A. Kozlov, \textbf{L. Czech}, B. Morel, D. Darriba, T. Flouri, and A. Stamatakis.
        EPA-ng: Massively Parallel Evolutionary Placement of Genetic Sequences.
        \textit{Systematic Biology}, 2018. \cite{Barbera2018}

    \item D. Bass, \textbf{L.Czech}, B. Williams, C. Berney, M. Dunthorn, F. Mahe, G. Torruella, G. Stentiford, and T. Williams.
        Clarifying the Relationships between Microsporidia and Cryptomycota.
        \textit{Journal of Eukaryotic Microbiology}, 2018.~\cite{Bass2018a}

    \item \textbf{L. Czech}, P. Barbera, and A. Stamatakis.
        Methods for Automatic Reference Trees and Multilevel Phylogenetic Placement.
        \textit{Bioinformatics}, 35(7):1151--1158, 2018. \cite{Czech2018-phat}

    \item \textbf{L. Czech} and A. Stamatakis.
        Scalable methods for analyzing and visualizing phylogenetic placement of metagenomic samples.
        \textit{PLOS ONE}, 14(5):e0217050, 2019. \cite{Czech2019-analysis}

    \item \todo{1KITE}
    \item \todo{long reads} \cite{Jamy2019a}  \todo{update reference}
    \item \todo{swarm 3?}
    \item \todo{genesis and gappa} \cite{Czech2019-genesis-gappa}
    \item \todo{dinos}
\end{enumerate}

\clearpage
